\textbf{Course: Física Teórica I (CDP7600)} \\[0.2cm] {\Large \textbf{Project}: \hspace{0.2cm} \textit{Fibration building blocks of information-processing networks} }

### 2019-10-31 Manuscript/main.tex
final art

Fibration building blocks of a information flow network represent the sets of nodes that are symmetric with respect to the processing of information, that is, the sets of nodes that process equivalent information. Here, we have reproduced important results concerning the identification and classification of the fibration building blocks of directed networks, constructed from real network data. More specifically, using the transcriptional regulatory network data of the \textit{Escherichia Coli} bacteria, we quantify the clusters of nodes that synchronously process equivalent information and then we classify these clusters, called network fibers, based on its specific topological features. This way, in order to consistently present the obtained results, in this report we first give a brief description of the theory concerning the graph fibration morphism and its main definitions related to information flow symmetries. Next, we detail the methods adopted to correctly identify and classify the network fibers. More specifically, to establish an optimal framework, I show the implementation details of the Minimal Balanced Coloring algorithm used to find the corresponding fibers in the network, presenting a slightly improvement for the algorithm complexity and its implementation. At last, showing the proper methods for data preparation, I describe the results obtained concerning the fiber statistics for the specific case of the \textit{Escherichia Coli} regulatory network, to properly compare with the recent results presented at \textit{Morone et. al.} (2019).

@@

In order to identify the correct distribution of network fibers we have to find a efficient procedure to split the network in different disjoint partitions, in which all nodes from a same partition should receive information from the same other partitions. To do that, we treat our problem as the same of find the coarsest relational refinement partitioning for the given network. In this section, we detail the optimal algorithm used to find this coarsest partitioning in a context of graph fibrations and show the relative simple implementation of this method.

@@

The algorithm used in this project is a slightly modified version of the algorithm presented by Paige and Tarjan \cite{Tarjan1987}, \ having a time complexity of $O(M\log N)$, being very efficient for sparse networks. This algorithm has the same runtime order than the algorithm from Cardon and Crushemore \cite{Cardon1982}, but it has a simpler implementation and smaller prefactors, exhibiting a better approach to our problem. Even though the algorithm applies for general situations, here we give the details of the algorithm to its application for a network, so we introduce the necessary definitions within this context.

A network $G$ is completely defined as $G(V, E)$ by the sets of nodes $V$ and of the connections $E$. The network has then $N = |V|$ nodes that can have connections with one another defined by the set $E$ containing $M = |E|$ ordered pair of nodes, denoting the directed connections between the network nodes. Over the network, we can define a graph partition $P$ of $V$ as a set of pairwise disjoints subsets of $V$ whose union is all $V$
\begin{equation}
	P = \bigcup_{i} P_i
\end{equation}
where $P_i$ are the elements of the partition $P$, in which we call it blocks. Within this context, if we consider an additional graph partition $Q$ that has the property that all the block of $Q$ is contained in a block of $P$, we say that $Q$ is a refinement partition of $P$.

Considering a block $B$ in $P$, we say that the block $B$ is stable with respect with a set $S$ if or all elements of $B$ connects an element of $S$ or none element of $B$ points to any element of $S$. By that we define that a graph partition $P$ of $V$ is stable if it is stable with respect with its own blocks. Having defined that, the graph coarsest partition problem is that of finding, for a given set of connections $E$ and initial partition $P$ over $V$, the coarsest stable refinement of $P$.

Considering all that, for the proper identification of the fibers on the network we have to construct a stable graph partition that is equivalent to the coarsest (minimal number of blocks) refinement of the network with respect to the information flow passing through each node. For that, we need to extend the concept of stability over the partition for one that accounts for the information received by each block of the partition. Thus, in order to identify the clusters with isomorphic input-tree for all their nodes, we require that the partition should be not only stable but input-tree stable. That means that for a subset $S \subseteq$ $V$, a graph partition $P$ over the network $G(V, E)$ is input-tree stable with respect to $S$ if for all the block $B \in P$, the equality
\begin{equation}
	| E^{-1}(\{x\}) \cap S | = | E^{-1}(\{y\}) \cap S |
\end{equation}
is satisfied for all the elements $x$, $y \in B$.

This way, we can benefit from the stability properties \cite{Tarjan1987} to construct a refinement algorithm step that can obtains, through a finite number of steps, a input-tree stable partition from an initial unstable partitioning of the network. With this, given a subset $S \subseteq V$, the refinement step has the effect to refine the current partition, unstable with respect to $S$, by replacing it for a new partition now stable with $S$. For that, we define a split function \textit{I-split($S$, $P$)} that receive as input the current partition and a set $S$ and results as output a new input-tree stable partition with respect to $S$. That function benefits from two major properties of stability (and input-tree stability): the stability inheritance by refinement and by union of sets. Since this refinement inheritance, a given set $S$ can be used only once by the function \textit{I-split}, for which after all other refinement will be stable with respect to $S$. Moreover, since stability is inherited under union of sets $S$, after sets are used in \textit{I-split}, their union cannot be used for the function. With all this considered, the essence of the algorithm can be stated as

\begin{quotation}
	Find a set $S$ in which the current partition $P$ is input-tree unstable. Replace $P$ by the output of \textit{I-split($S$,$P$)}. Guarantees that the set $S$ or unions of used sets never be used again.
\end{quotation}

Since the finest partitioning possible is the one in which every node is itself a block, the refinement step can be proceeded at most $N-1$ times, guaranting that the algorithm terminates with the correct answer, since stability is inherited by the refinement process. However, to guarantees that the algorithm has a optimal runtime we have to find a efficient way to select the appropriate sets to the refinement step, without choosing repeated sets. Fortanately, this can be easily done for the construction of a input-tree stable partition.

Given a set $S$ of nodes from the network $G$ and a given input-tree unstable graph partition $Q$, the blocks $B \in Q$ that are input-tree unstable with respect to $S$ can be splitted in several blocks $B_j$ to have a stable input-tree for $S$. Then, each splitted block will have the property defined by
\begin{equation}
	B_j = \{ x \in B : | E^{-1}(\{x\}) \cap S | = j \}
\end{equation}
where the number of splitted block must be larger than one to a proper split process take place. Then, for each unstable block $B \in Q$ all the splitted block, except the largest one, can be put a queue to be used ahead in the algorithm as a refinement set. This ensures that none repeated sets or union of repeated sets can be used during the algorithm exeecution.

Having said all that, the complete algorithm to find the correct network fibers of a network $G(V, E)$ consists in initializing a graph partition $Q$ over all nodes from $V$ except the nodes that do not receive information from any other node, in which each one of these will be defined as isolated fibers already in the beginning of the algorithm. The partition $Q$ is defined as one block containing all the other nodes in the network. The algorithm maintains a queue $L$ of possible refinement sets, initially containing the single block of $Q$ and all the isolated blocks defined at the beginning. Then, we proceed as
\begin{quotation}
	Remove from $L$ its first set $S$. Replace $Q$ by the \textit{I-split($S$, $Q$)}. Whenever a block $B \in Q$ splits into two or more nonempty blocks, add all but the largest to the back of $L$.
\end{quotation}
And this process is repeated until the queue $L$ is empty. At this point, the resulted partition $Q$ represent the coarsest input-tree stable partitioning of the network $G(V, E)$, where each block represents a network fiber with all its nodes having isomorphic input-trees.

>>>>

Fibration building blocks of information-processing networks represent the sets of nodes that are symmetrical with respect to the processing of input signals received from the rest of the network, defining this way the classes of nodes that process equivalent information. In this report, I reproduced relevant results concerning the identification and classification of the fibration building blocks of directed networks, constructed from real systems data. More specifically, using the transcriptional regulatory network data from the \textit{Escherichia Coli} bacteria, I quantify the clusters of nodes in the network that synchronously process equivalent information, and then I classify these clusters, called fibers, based on its specific topological features. Here, to consistently present the obtained results, I briefly introduce the theory concerning the graph fibration morphism and its main definitions related to information processing symmetries. Then, I detail the computational methods adopted to the identification of the network fibers, presenting a slightly different algorithmic approach for the problem in hand, based on the general relational refinement partitioning algorithms described by Paige and Tarjan in 1987. After defining the implementation details of the graph refinement algorithm I lastly apply the method for trial network examples and then for the constructed genetic regulatory network of \textit{Escherichia Coli}, noticing a visible discrepancy between the results obtained by the methods described in this report and the results exhibited in the most recent literature given at \textit{Morone et. al.} (2019). The results showed in this report, concerning the fibers distribution over the network, are tested and verified to support the assertion that the algorithm used captures the correct minimal fibration.

@@

To identify the correct distribution of fibers over the network it is necessary to define an efficient procedure to split the nodes into different disjoint partitions, from which all nodes inside the same partition must receive equivalent information from all other partitions. To do that, we treat our problem as the same as finding the coarsest relational refinement partitioning of a set of elements that have a binary relation between them. Since a network is completely defined by a set of node elements and a set of edges that comprehend all the binary relations, this approach can be used. Therefore, in this section, I detail the optimal algorithm used to find this coarsest partitioning in the context of graph fibrations and show the relatively simple implementation of this method.

The algorithm used in this project is a slightly modified version of the algorithm presented by Paige and Tarjan \cite{Tarjan1987}, \ having a runtime complexity of $\mathcal{O}(M\log N)$, where $M$ and $N$ are, respectively, the number of edges and nodes in the network, being almost linear with the size of the network. It is important to note that this algorithm has the same runtime order than the algorithm from Cardon and Crushemore \cite{Cardon1982}, in which the algorithm of minimal fibration from \cite{fibration2019} is based. However, the algorithm presented by Paige and Tarjan has a simpler implementation and smaller prefactors, exhibiting a better approach to our problem. Even though the algorithm has wider general applications, here I give the details of the algorithm implementation to the application for a network, so that I introduce the necessary definitions within this context.

@@

A network $G(V, E)$ is completely defined by the sets of the nodes elements $V$ and the set of the edges $E$. The network has $N = |V|$ nodes that can have connections with one another defined by the set $E$, which contains $M = |E|$ ordered pair of nodes, denoting the directed links between the network nodes. Considering the network, we can define a graph partition $\bar{P}$ over $V$ as a set of pairwise disjoints subsets of $V$ whose union is all $V$, that is
\begin{equation}
	\bar{P} = \bigcup_{i} P_i
\end{equation}
where $P_i$ are the elements of the partition $\bar{P}$, called blocks. Considering this, if we take an additional graph partition $\bar{Q}$ that has the property that all of its blocks are contained in a block of $\bar{P}$, we say that $\bar{Q}$ is a refinement partition of $\bar{P}$.

Moreover, for a block $B \in \bar{P}$, we say that the block $B$ is stable with respect with a set $S$ if either all elements of $B$ connects with an element of $S$ or none element of $B$ points to any element of $S$. The stability partition concept is central to the problem of relational coarsest refinement partitioning and can be easily extended to graph problems, where the coarsest graph partition problem is that of finding, for a given set of connections $E$ and an initial partition $\bar{P}$ over $V$, the coarsest stable refinement of $\bar{P}$, i. e., the minimal number of disjoint blocks subsets of $V$ that forms a stable refinement of $\bar{P}$.

Considering the stability of partitions, for a proper identification of the network fibers as defined in the above section we ought to construct a stable graph partition that is equivalent to the coarsest refinement of the network (minimal number of blocks) with respect to the information processing of each node. For that goal, we need to extend the concept of stability over partitions for the case that accounts for the information processed by each block of the partition. Therefore, to identify the group of nodes with isomorphic input-tree, we require that the partition should be not only stable but \textbf{input-tree stable}. That means that for a subset $S \subseteq$ $V$, a graph partition $\bar{P}$ over the network $G(V, E)$ is input-tree stable with respect to $S$ if for all the block $B \in \bar{P}$, the following equality is satisfied for all the elements $x$, $y \in B$:
\begin{equation}
	| E^{-1}(\{x\}) \cap S | = | E^{-1}(\{y\}) \cap S |
\end{equation}
where $E(\{x\})$ and $E^{-1}(\{x\})$ represents, respectively, the set of nodes that directly receives information from $x$, and the set of nodes that sends information, via a direct link, to $x$.

This way, we can benefit from the stability properties \cite{Tarjan1987} to construct a refinement algorithm step that can achieve, through a finite number of steps, a input-tree stable partition from an initial unstable partitioning of the network. This way, given a subset $S \subseteq V$, the refinement step has the effect to refine the current partition, input-tree unstable with respect to $S$, by replacing it for a new partition, now input-tree stable for $S$. With that objective, we define a split function \textit{I-split($S$, $\bar{P}$)} that receive as input the current partition and a set $S$ and returns as output a new input-tree stable partition with respect to $S$. Favorably, that function benefits from two major properties of stability (and input-tree stability): the stability inheritance by refinement and by union of sets. By reason of this refinement inheritance, a given set $S$ can be used only once by the function \textit{I-split}, guaranting that the partition, after all other refinement steps, will maintain as stable with respect to $S$. Moreover, since stability is inherited under union of sets $S$, after sets are used in \textit{I-split}, their union cannot be used for the function. Considering all these properties, the essence of the refinement algorithm can be stated as

\begin{quotation}
	\textbf{Refinement Algorithm}: Find a set $S$ in which the current partition $\bar{P}$ is input-tree unstable. Replace $\bar{P}$ by the output of \textit{I-split($S$,$\bar{P}$)}. Guarantees that the set $S$ or unions of used sets never be used again.
\end{quotation}

Since the finest partitioning possible is the one in which every node is itself a block, the refinement step may be proceeded at most $N-1$ times, guaranting that the algorithm terminates with the correct answer, since stability is inherited by the refinement process. However, to guarantees that the algorithm has a optimal runtime we have to find a efficient way to select the appropriate sets to the refinement step, without choosing repeated ones. Fortunately, this can be easily done for the construction of a input-tree stable partition.

Given a set $S$ of nodes from the network $G(V, E)$ and a given input-tree unstable graph partition $\bar{P}$, the blocks $B \in \bar{P}$ that are input-tree unstable with respect to $S$ can be splitted into several blocks $B_j$ that will be stable input-tree with respect to $S$. Then, each splitted block will have the property defined by
\begin{equation}
	B_j = \{ x \in B : | E^{-1}(\{x\}) \cap S | = j \}
\end{equation}
where the number of splitted block must be larger than one to a proper splitting process take place. Then, for each unstable block $B \in \bar{P}$ all the splitted block, except the largest one, can be put at the back of a queue to be used ahead in the algorithm as a refinement set. This ensures that none repeated sets or union of repeated sets can be used during the algorithm exeecution.

Finally, having said all that, the complete algorithm to find the correct network fibers of a network $G(V, E)$ consists in initializing a graph partition $\bar{P}$ over all nodes from $V$ except the nodes that do not receive information from any other node, in which each one of these will be defined as isolated fibers already in the beginning of the algorithm. The partition $\bar{P}$ is defined as one block containing all the other nodes in the network. The algorithm maintains a queue $L$ of possible refinement sets, initially containing the single block of $\bar{Q}$ and all the isolated blocks defined at the beginning. Then, we proceed as
\begin{quotation}
	\textbf{Coarsest Refinement Algorithm}: Remove from $L$ its first set $S$. Replace $Q$ by the \textit{I-split($S$, $Q$)}. Whenever a block $B \in Q$ splits into two or more nonempty blocks, add all but the largest to the back of $L$.
\end{quotation}
And this process is repeated until the queue $L$ is empty. At this point, the resulted partition $\bar{P}$ represents the coarsest input-tree stable partitioning of the network $G(V, E)$, where each block represents a network fiber with all its nodes having isomorphic input-trees.
